A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)

### 2019-12-09 Simulation/Writeup/NoteConvergence.Rnw

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Hanne Ida Oberman, BSc.\\
  Methodology and Statistics for the Behavioural, Biomedical and Social Sciences\\
  Department of Methodology and Statistics\\
  Faculty of Social and Behavioral Sciences\\
  Utrecht University\\
  Heidelberglaan~15\\
  3500 Utrecht, The Netherlands\\
  E-mail: \email{h.i.oberman@uu.nl} % \\
  URL: \url{https://hanneoberman.github.io/}
}

@@

Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} \textbf{[Expand or remove Rhat here! Refer to future section?]}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this is appropriate, and numerical is not]}. Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.

@@

The aim is to evaluate whether the imputation procedure has converged. The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.

@@

To investigate convergence properties of the \pkg{mice} algorithm, different lenghts of imputation chains are considered. this simulation study considers 100 simulation conditions. Each condition has a different maximum number of iterations, varying between one and one-hundred iterations. In this study, 1000 simulation runs are performed for each of the 100 simulation conditions. The simulation set-up consists of several steps, summarized in the pseudo-code below. The complete \proglang{R} script of the simulation study is available on Github.

@@

\subsection{Data}
The simulated dataset is a finite population of $N=1000$. The data are simulated to solve a multiple linear regression complete data problem. The quantity of scientific interest is the estimated regression coefficient of predictor $X$ on outcome variable $Y$. The linear regression model is $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$, where $Y$ is the dependent variable, $X$ is an independent variable, and $Z_1$ and $Z_2$ are covariates. The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$. Outcome variable $Y$ is deduced from the predictor variables as $Y =  2 \times X + 0.5 \times Z_1 - 1 \times Z_2 + \epsilon$.

@@

\subsection{Amputation}
The complete data is the starting point for all simulations. In each repetition of the simulation study, the complete data is 'amputed' once. That is, the \pkg{mice} function \fct{ampute} in \proglang{R} is used to impose an MCAR missingness mechanism upon the data. The missingness is univariate, and the probability to be missing is the same for all variables, namely 20\%. This leaves 20\% of the rows completely observed. The resulting amputed data is equal for all simulation conditions in the same repetition.%"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9. Note that these values indicate the sampled proportion of incomplete cases in Y" \cite{schou18}.

@@

\subsection{Simulation diagnostics}
The simulation diagnostics are as recommended by \cite{buur18}. These comprise of average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations. %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd}. For now, this is outside of the scope of this study. % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

@@

\textbf{$\widehat{R}$}. Steep decline up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that. Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations considered in this simulation study (see online appendix for the full table of results).

\textbf{Auto-correlation.} Within the 100 iterations considered in this study, AC decreases steeply, then slowly increases. Lowest AC observed is \Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)} at four iterations. There appears to be a trend towards zero from four iterations upwards. Auto-correlation does not plateau completely. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.

@@

\textbf{add loess line to plots to see when it becomes a flat line.}

@@

\textbf{Summary.}
This note illustrates that conventional convergence diagnostics behave differently on MI data than other MCMC methods like Gibbs samplers in Bayesian analyses. The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) is too stringent for MI data.

From the simulation diagnostics, it appears that as little as two iterations could be sufficient to draw valid inferences. Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate deviant behavior up-to twenty or even forty iterations.

\textbf{R hat below 1.}
Moreover, $\widehat{R}$ could theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ). How could $\widehat{R}$ smaller than 1 occur? The number of simulations is smaller than in `regular' MCMC processes [explain that fewer iterations is an advantage of MI, not a disadvantage compared to MCMC]. Therefore, the `$(n-1/n)$' [add equation number] correction factor can influence the estimated potential scale reduction factor. This downwards bias is in the opposite direction than expected: ``The mixture-of-sequences variance, V, should stabilize as a function of n. (Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)'' \cite[p~438]{broo98}.

@@

The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the \pkg{mice}~\Sexpr{paste(packageVersion("mice"), collapse = ".")} package. \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

>>>>

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Hanne Ida Oberman, BSc.\\
  Methodology and Statistics for the Behavioural, Biomedical and Social Sciences\\
  Department of Methodology and Statistics\\
  Faculty of Social and Behavioral Sciences\\
  Utrecht University\\
  Heidelberglaan~15, 3500 Utrecht, The Netherlands\\
  E-mail: \email{h.i.oberman@uu.nl} % \\
  URL: \url{https://hanneoberman.github.io/}
}

@@

Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} \textbf{[Expand or remove Rhat here! Refer to future section?]}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this can still be appropriate, and numerical is not]}. Visually assessing convergence is not only challenging to the untrained eye, it might also be futile.

@@

The aim is to evaluate whether the imputation procedure has converged. The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data. \textbf{Or if there is a discrepancy between the visual and numeric inspection.}

@@

To investigate convergence properties of the \pkg{mice} algorithm, different lenghts of imputation chains are considered. This simulation study considers 100 simulation conditions. Each condition has a different maximum number of iterations, varying between one and one-hundred iterations. In this study, 1000 simulation runs are performed for each of the 100 simulation conditions. \textbf{where for one through one hundred the Rhat and ... are calcuated.} The simulation set-up consists of several steps, summarized in the pseudo-code below. The complete \proglang{R} script of the simulation study is available on Github.

@@

\subsection{Data}
The simulated dataset is a finite population of $N=1000$. The data are simulated to solve a multiple linear regression complete data problem. The quantity of scientific interest is the estimated regression coefficient of predictor $X$ on outcome variable $Y$. The linear regression model is $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$, where $Y$ is the dependent variable, $X$ is an independent variable, and $Z_1$ and $Z_2$ are covariates. The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$.

@@

Outcome variable $Y$ is subsequently calculated as $Y =  2X + .5Z_1 - Z_2 + \epsilon$.

\subsection{Amputation}
The complete data is the point of origin for all simulations. In each repetition of the simulation study, the complete data is `amputed' once. That is, the \pkg{mice} function \fct{ampute} in \proglang{R} is used to impose an MCAR missingness mechanism upon the data. The missingness is univariate, and the probability to be missing is the same for all variables, namely 20\%. This leaves 20\% of the rows completely observed. The resulting amputed data is equal for all simulation conditions in the same repetition.%"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9. Note that these values indicate the sampled proportion of incomplete cases in Y" \cite{schou18}.

@@

\subsection{Simulation diagnostics}
The simulation diagnostics are as recommended by \cite{buur18}. These comprise of average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations. %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.). For now, this is outside of the scope of this study. % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

@@

\textbf{$\widehat{R}$}. Steep decline up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that. Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations considered in this simulation study (see online appendix for the full table of results). \textbf{[REWRITE AS STORY, NOT JUST DATA.]}

\textbf{Auto-correlation.} Within the 100 iterations considered in this study, the AC decreases steeply, before a slow increase. The lowest AC observed is \Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)} at four iterations. There appears to be a trend towards zero from four iterations upwards. Auto-correlation does not plateau completely. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.

@@

\textbf{add loess line to plots to see when it becomes a flat line. also, give some implications for these measures.}

@@

\textbf{Summary.}
This note illustrates that conventional convergence diagnostics behave differently on MI data than other MCMC methods like Gibbs samplers in Bayesian analyses. The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) may be too stringent for MI data.

From the simulation diagnostics, it appears that as little as two iterations could be sufficient to draw \textbf{[valid inferences?]}. Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate that convergence may only be reached after twenty or even forty iterations.

\textbf{R hat below 1.}
$\widehat{R}$ could theoretically not be smaller than one, yet it occurred several times in this study (see online appendix XYZ). This can happen when the number of simulations is smaller than in `regular' MCMC processes [explain that fewer iterations is an advantage of MI, not a disadvantage compared to MCMC]. Therefore, the `$(n-1/n)$' [add equation number] correction factor can influence the estimated potential scale reduction factor. This downwards bias is in the opposite direction than expected: ``The mixture-of-sequences variance, $V$, should stabilize as a function of $n$. (Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)'' \cite[p~438]{broo98}.

@@

The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} \cite{R} with the \pkg{mice}~\Sexpr{paste(packageVersion("mice"), collapse = ".")} package \cite{mice}. \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

### 2019-12-09 Simulation/Writeup/NoteConvergence.tex

Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} \textbf{[Expand or remove Rhat here! Refer to future section?]}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this is appropriate, and numerical is not]}. Visually assessing convergence is not only difficult to the untrained eye, it might also be futile. The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}. Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.

@@

\textbf{$\widehat{R}$}. Steep decline up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that. Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations considered in this simulation study (see online appendix for the full table of results).

\textbf{Auto-correlation.} Within the 100 iterations considered in this study, AC decreases steeply, then slowly increases. Lowest AC observed is -0.735 at four iterations. There appears to be a trend towards zero from four iterations upwards. Auto-correlation does not plateau completely. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.

@@

The results in this paper were obtained using \proglang{R}~3.6.1 with the \pkg{mice}~3.6.0.9000 package. \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

>>>>

Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} \textbf{[Expand or remove Rhat here! Refer to future section?]}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this can still be appropriate, and numerical is not]}. Visually assessing convergence is not only challenging to the untrained eye, it might also be futile.

@@

\textbf{$\widehat{R}$}. Steep decline up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that. Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations considered in this simulation study (see online appendix for the full table of results). \textbf{[REWRITE AS STORY, NOT JUST DATA.]}

\textbf{Auto-correlation.} Within the 100 iterations considered in this study, the AC decreases steeply, before a slow increase. The lowest AC observed is -0.735 at four iterations. There appears to be a trend towards zero from four iterations upwards. Auto-correlation does not plateau completely. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.

@@

The results in this paper were obtained using \proglang{R}~3.6.1 \cite{R} with the \pkg{mice}~3.6.0.9000 package \cite{mice}. \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.
